% - \textbf{Category geometry} Within-model category directions are low-to-moderately correlated and show weak cluster quality (negative Silhouette), yet remain usable for targeted control (Sec. ~\ref {fig:steering_vector_cos_sim}).

% These signals support the hypothesis that categorical refusal tokens shape the representation space in a directionally meaningful way that is only partly aligned with pre-existing base-model features.

\subsection{Limitations and Future Work}

- \textbf{Evaluation fidelity} LLM-as-judge metrics depend on the judge’s criteria and calibration. While convenient and widely used, they may not perfectly capture user utility.
- \textbf{Coverage} We focus on a single steering site (layer 9, resid\_post) and five refusal categories. Other sites/layers and broader category taxonomies could strengthen generality.
- \textbf{Metrics availability} Some base-model token-level metrics are unavailable (N/A), limiting direct like-for-like comparisons.
- \textbf{Open analyses} We did not include inference-time steering ablations or SAE-based steering in the final tables. Integrating these and reporting confidence intervals for all tasks are priorities for follow-up.

\paragraph{Outlook} The results indicate a promising path: targeted safety gains with emerging robustness to adversarially benign prompts. Future work should pursue multi-objective fine-tuning to mitigate capability regressions, richer mechanistic probes (e.g., causal tracing), and calibrated control policies that adaptively modulate category-specific directions at inference time.
